\documentclass[12pt]{article}
\usepackage[margin=0.5 in]{geometry}
\usepackage{color}
\pagenumbering{gobble}
\begin{document}
\title{Project Plan\\ {Automata and Numeration Systems} \\Spring 2019}
\author{Faculty Mentors: {Philipp Hieronymi and Erik Walsberg} \\
	Project Leaders: {Eion Blanchard and Alexi Block Gorman}\\
	IGL Scholars: {Dun Ma, Steve O'Brien, Reed Oei, Mihika Poddar, Dagoberto Saenz }}
\date{\today}
\maketitle
\section{Project Goals}
\begin{itemize} 
\item Validate the general addition and verification automaton for Ostrowski Numeration System. 
\item Implement a generator for infinite balanced words. 
\item Give proofs for critical exponents of infinite balanced words of large alphabets. 
\item Write up the results and produce a paper. 
\item If possible, implement a more general automaton generator for Ostrowski Numeration System. 
\end{itemize}

\section{Responsibilities}
\begin{itemize}
\item Every group member is responsible for reading the required texts and papers for background knowledge.
\item Every group member should be familiar with running predicates in walnut. 
\item We will discuss the responsibility for theorem proving when everyone is familiar with the project background.
\item Coding responsibilities will be discussed when there is a need to do so. 

\end{itemize}

\section{Meeting Times}
\begin{itemize}
	\item Everyone will meet weekly at 3-4pm on Mondays.
	\item Graduate mentors and undergraduates will meet weekly at 5-6pm on Wednesdays.
	\item Undergraduate students will meet over the weekend if needed. 
\end{itemize}
\end{document}
